---
title: "A Guide to Lakhesis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Guide to Lakhesis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lakhesis)
```

This vignette provides a guide to the R package `lakhesis`. Please refer to the paper, "Lakhesis: Consensus Seriation via Iterative Regression of Partial Rankings for Binary Data," for details.

## Introduction

Seriation can be accomplised using a variety of methods. The main goals of the `lakhesis` package, which focus on binary matrices (i.e., presence/absence, 0/1), are to provide:

1. an interactive graphical interface for investigators to select their own well seriated data and perform seriation via correspondence analysis;
2. the means to establish a single consensus seriation from multiple, partial seriations via a critical coefficient.

While there are already a number of approaches and tools to perform seriations, `lakhesis` uses a technique of fitting CA scores to those of a curve produced by an ideal, well seriated "reference matrix" via a Procrustes method (center, scale, and rotate).

Given that well seriated matrices are not a given for every data set, it is expedient for users to select their own seriations quickly via an interactive graphical interface, rather than slicing matrices and re-running commands via the console. Investigators can select multiple seriations, moreover, which can then be harmonized into a single seriated matrix.

The `lakhesis` package, named for the ancient Greek goddess who measured the thread of one's life, provides this functionality, to explore and select well seriated sequences as "strands" that are then "threaded" together into a single consensus seriation. The package can either be installed from [CRAN](https://CRAN.R-project.org/package=lakhesis), or from [GitHub](https://github.com/scollinselliott/lakhesis), the latter of which hosts the most recent development version. To install from GitHub, the library `devtools` is needed:
```{r, eval = FALSE}
library(devtools)
install_github("scollinselliott/lakhesis") 
```

This vignette showcases the primary functions of the package side-by-side with the Lakhesis Calculator, a `shiny` app whose interface offers the means to perform and critically examine seriated sequences from an input. Command-line functions to prepare 

## Getting Started: Formatting Data

The calculator is designed to import `csv` files directly from the user's file directory. The `csv` file must be formatted in a "long" formamt where row and column incidences are listed in rows, pair by pair. To convert data that is already in the form of a matrix, the `im_long()` function can used.  The `"quattrofontanili"` data can serve as an example, which comprises a matrix of 81 rows and 82 columns, recording the incidence of a particular artifact-type (column) in a tomb (row) from Early Iron Age necropoleis in Southern Etruria (see `?quattrofontanili`).
```{r}
library(lakhesis)
data("quattrofontanili")
qf <- im_long(quattrofontanili)
```
The `data.frame` `qf` can then be exported to a `csv` file using the `write.table()` function, in order to avoid writing row and column headers:
```{r, eval = FALSE}
write.table(qf, file = "qf.csv", row.names = FALSE, col.names = FALSE, sep = ",")
```
The file `qf.csv` is then ready to be uploaded into the Lakhesis Calculator.

### Pooling Observations

If one has multiple data sets to pool together, it is simply a matter of added additional rows to the `csv` file. The `im_merge()` function assists with creating a single incidence matrix from two different incidence matrices, overriding any 0 values with 1s in the case of shared row and/or column elements. The merged matrix can then be converted using `im_long()` and exported, as above.

```{r, eval = FALSE}
qf1 <- quattrofontanili[1:20, 1:40]
qf1 <- qf1[rowSums(qf1) != 0, colSums(qf1) != 0]

qf2 <- quattrofontanili[30:50, 20:60]
qf2 <- qf2[rowSums(qf2) != 0, colSums(qf2) != 0]

im_merge(qf1, qf2)
```

## Procrustes-Fit Correspondence Analysis

The method developed here uses correspondence analysis (CA), a popular method used for seriation above all in the form of detrended correspondence analysis (DCA). Unlike DCA, this package seriates observations according to a "reference curve," since the principal scores of well seriated rows and columns will have the form of an arch or horseshoe in the CA biplot. A well-seriated "reference matrix" of the same size is also plotted with CA, and the principal scores of the data are fit to the scores of the reference matrix using a Procrustes method (center, scale, rotate). The axes are properly labelled "Procrustes1" and "Procrustes2" since they no longer represent the principal scores of the data. Procrustes fitting is done without "landmark points" by minimizing the distance of scores to those of the reference curve in a one-to-many mapping.

This is accomplished with the function `ca_procrustes()`, which takes an incidence matrix as an input and returns an object `procrustean` that contains the reference and data scores. The reference matrix is generated internal to `ca_procrustes()` using the `im_ref()` function. The `plot` function automatically illustrates the fit of the points to the reference curve:
```{r,  fig.align = 'center'}
data("quattrofontanili")
qf_caproc <- ca_procrustes(quattrofontanili)
class(qf_caproc)
print(qf_caproc)
plot(qf_caproc)
```

In order to determine a seriation for both the row and column points, they are projected onto the quadratic model which fits the reference points. Then, their ranking is taken from one end of the curve to another. The function `ca_procrustes_ser()` takes an incidence matrix as input, and produces as output a `strand` object, which comprises a `list` that includes data regarding the row and column score points, distance, and rankings, as well as the seriated matrix according those rankings.
```{r}
qf_ser <- ca_procrustes_ser(quattrofontanili)
class(qf_ser)
print(qf_ser)
summary(qf_ser)
```

Three different plot types are available for a `strand` object, set in the `plot()` function via the additional `display` argument:

* `"ca"` produces a Procrustes-fit CA plot as above, with axes $\text{Procrustes1}$ and $\text{Procrustes2}$. This is the default setting.
* `"ref"` produces a plot showing the distance of each row and column point from the reference curve. The indices of the reference curve are shown on the $x$ axis while the distance of each point from the curve is shown on the $y$ axis.
* `"im_seriated"` produces a matrix plot of the seriated incidence matrix, showing the measure of concentration $\kappa$ (on which see below).

```{r, fig.align = 'center'}
plot(qf_ser, display = "ca")
plot(qf_ser, display = "ref")
plot(qf_ser, display = "im_seriated")
```

The resulting incidence matrix may be extracted from the `strand` by calling `$im_seriated`:

```{r}
quattrofontanili.caproc <- qf_ser$im_seriated
```

## Lakhesis Analysis

In the process of exploring matrices for seriated sequences, an investigator is liable to come across many sequences which appear to be well seriated. The question then arises as to how to harmonize or reconcile several partial, seriated sequences, some or many of which may disagree. How to reconcile these is non-trivial, analogous to other problems in e.g. ranked-choice or preference voting, but it is not strictly identical.

The strategy adopted here, called a "Lakhesis" method, is to iterate regressions of partial sequences onto best-fitting sequence in an agglomerative fashion. The procedure consists of the followings steps:

1. Randomly choose two strands (partial sequences) which have at least 4 joint elements.
    a. Remove all elements for which both strands have `NA` values.
    b. Temporarily suppress the elements for which at least one strand has an `NA` value, leaving just elements which have rankings in both strands.
2. Perform eigendecomposition (unscaled PCA) on the rankings of the stands to find a best fitting line with respect to both rankings.
3. Restore the elements which were temporarily suppressed in either strand, and perform simple linear regression separately for either strand onto that best fitting line.
    a. If the same element in either strand is regressed onto different points along the best fitting line, the mean of their regression is used.
    b. The elements are re-ranked along the regression line, providing a single "harmonized" ranking from the two strands.
4. Take this harmonized ranking as the next strand and choose another strand at random that has at least 4 joint elements with the harmonized ranking, and repeat the above Steps 1a - 3b.
5. Steps 1a - 4 are repeated until all strands have been harmonized, producing a single ranking.

Because the order in which strands have been randomly selected will bear on the estimation of their single, harmonized seriation, these steps are repeated for a number of iterations. The resulting set of harmonized rankings are then subjected to a final eigendecomposition, with points projected onto the first principal component axis to determine their consensus ranking. The use of PCA/linear regression has its virtue in that partial rankings which may be identical in their relative order but which may run in reverse sequence, will be regressed into the same order but merely with an opposite eigenvector or negative slope.

### The `lakhesize()` Function

This task is performed for a set of strands by the `lakhesize()` function, which takes as its input a `strands` object (i.e., a `list` of individual `strand`s produced by `ca_procrustes_ser()`).

The `lakhesize()` function has the following options:

* `iter` The number of iterations to produce a consensus seriation (the default exploratory option is 100).
* `sim` The number of consensus seriations produced via simulation (the default exploratory option is 1).
* `pbar` Turns the progress bar on or off. Default is `TRUE`.

The default setting of 100 iterations using 1 simulation run in the Lakhesis Calculator provides an approximation for an optimal consensus seriation, determined by the selection of strands by the investigator. Adjusting `iter` and `sim` afterwards can be used to determine if there is an even more optimal seriation.

A predefined `strands` object is included in the data of the package to illustrate functionality of the `lakhesize()` function. The `qfStrands` object contains three strands from the `quattrofontanili` data:

```{r}
data("qfStrands")
print(qfStrands)

set.seed(42)

# the default used in the Lakhesis Calculator
L.exp <- lakhesize(qfStrands, pbar = FALSE)

# employing more simulations to detect a more optimal solution
L.opt <- lakhesize(qfStrands, iter = 20, sim = 20, pbar = FALSE)

summary(L.exp)
```

The output is contained in a `lakhesis` object, which includes the following objects as described in the `summary()` above:

* seriated sequences of the row and column elements according to the consensus seriation of the input strands
* PCA results of the final consensus seriation
* the seriated incidence matrix
* coefficients of concentration and agreement (on which see below)

## Critical Measures

The `lakhesis` package employs three methods to critically evaluate a seriated incidence matrix.

### Concentration ($\kappa$)

As discussed in the paper at the start of this vignette, there are a number of optimality measures for seriated matrices. In this package, a measure of concentration has been defined, $\kappa$, which is a loss measure with a minimum of 1. Concentration is based on the number of non-contiguous values of 1 attested in the seriated matrix across both columns and rows. The intervals of the index from the lowest-ranked 1 and that of the highest ranked 1 is summed for each row and column, and then divided by twice total sum of the matrix. A seriated matrix that has perfect concentration will thus have $\kappa = 1$, and should there occur 0s in between values of 1, the measure of $\kappa$ will increase. This is performed using the function `conc_kappa()`, whose documentation contains more information, which takes as input an incidence matrix.

First, we can create two objects of the seriated incidence matrices, one from the results of Lakhesis analysis using `"exploratory"` option (`L.exp`) and the other using the `"optimize"` option:
```{r}
quattrofontanili.L.exp <- L.exp$im_seriated
quattrofontanili.L.opt <- L.opt$im_seriated
```
Then, we can assess the concentration measure $\kappa$ for each seriated matrix:
```{r}
# the original seriation
conc_kappa(quattrofontanili)
# using procrustes-fit ca
conc_kappa(quattrofontanili.caproc)
# default lakhesis
conc_kappa(quattrofontanili.L.exp)
# optimized lakhesis
conc_kappa(quattrofontanili.L.opt)
```
Lower concentration measures inidicate more optimal seriations. As can be seen, the optimized Lakhesis analysis has arrived at a slightly more optimal seriation according to the concentration principle. The number of iterations and simulation runs can be based on the size of the incidence matrix. Here, 20 iterations with 20 simulation runs is more than sufficient to explore all possible consensus serations for 3 strands.

### Agreement ($\rho_r^2 \rho_c^2$)

The agreement of each strand's ranking with that of the resulting consensus seriation also serves as a useful diagnostic to determine if any one strand is in severe disagreement and should be removed before re-running Lakhesis. Agreement is defined as the product of Spearman's rank correlation coefficient squared for the row and column elements between one seriated matrix and another. In the `lakhesize()` function, agreement is computed for each strand's seriation in connection with the consensus seriation, using the function `spearman_sq()`, which removes `NA` values automatically and takes a numerical input of the rankings.

### Deviance Test for Goodness-of-Fit

The use of a quadratic-logistic model for determining seriations has a long history in ecology, but as a measure for optimality it poses a challenge in that an optimally concentrated seriation would elicit `NA` results for a logistic regression, owing to the perfect separation of 0/1 values in the columns and rows.

While a deviance-based test for goodness-of-fit is often performed for (nested) models upon the same data, here it is used as an exploratory method to evaluate which specific row and column elements might not fit well, and which one may wish to exclude when prior to performing correspondence analysis. The function `element_eval()` takes a seriated matrix as its input and will return a `list` of two `data.frame`s sorted from highest to lowest in terms of the resulting $p$ value from a deviance test using the quadratic-logistic model. Any rows or columns which exhibit perfect separation are first assigned a result of `NA`. Then, the deviance test is run on the remaining rows and columns. Elements which contain higher $p$ values will have a less optimal fit. 
```{r}
qf_eval <- element_eval(quattrofontanili)
head(qf_eval$RowFit)
head(qf_eval$ColFit)
```

## Plotting Strands

Plotting the results of `lakhesize()` can be accomplished by stipulating different options for the `display` argument in a `plot()` function, taking a `lakhesis` object as input.

* `"im_seriated"` will produce a matrix plot of the consensus seriation, also displaying the concentration measure $\kappa$. This is the default.
* `"rowPCA"` and `"colPCA"` will produce a biplot of the row and column element PCA. The closer the eigenvectors lie along the same trajectory, the more the strands will agree with one aother.
* `"agreement"` will produce a bar graph showing the measures of agreement of each strand with the consensus. Higher values indicate better agreement.
* `"concentration"` will produce a bar graph of each strand's concentration measure $\kappa$. Lower values indicate more optimal seriations.

```{r, fig.align = 'center'}
plot(L.exp, display = "im_seriated") 
plot(L.exp, display = "agreement") 
plot(L.exp, display = "concentration") 
```

These plots are incorporated into the Lakhesis Calculator, and are generated automatically.

## Lakhesis Calculator

The commands above can all be used in the console, and, as recommended, performing `lakhesize()` with more simulations will help to ensure that the most optimal consensus seriation will be identified. That said the functionality of the `lakhesis` package is best articulated via the interactive Lakhesis Calculator, written using `shiny`, which is initialized by the function `LC()`:
```{r, eval = FALSE}
LC()
```
This executes a `shinyApp()` function that will launch the calculator in the system browser. In the browser window, a sidebar is located on the left, while four panels are visible in the main dashboard:

* Seriation Explorer
* Consensus Seriation
* Criteria
* Modify

The calculator initializes with an ideally seriated reference matrix $20 \times 20$ in size, with row elements labeled "R" and column elements labeled "C." This reference matrix is produced using the function `im.ref()`:
```{r, eval = FALSE}
im.ref( matrix(NA, nrow = 20, ncol = 20) )
```

### Uploading Data

Before selecting the dataset to upload, choose whether or not the calculator will automatically remove rows and columns which are attested only once. This is done in the top left sidebar with the radio button under **Before Uploading**. The option **Use All Data** will use the full data when generating the incidence matrix within the calculator. The option **Remove Hapax** will remove all rows and columns which have only one attestation, since these instances do not establish connective relationships with other rows and columns. This may be helpful if one has a large matrix.

To upload data, click the **Browse...** button at the top of the sidebar under the **Choose CSV file** heading, and select the `qf.csv` file exported earlier.

### Identifying and Selecting Seriated Sequences

The plot in the upper left panel, Seriation Explorer, will automatically adjust from the initial default to display the correspondence analysis (CA) plot which has been Procrustes-fit to its reference curve. The Seriation Explorer contains two tabs, each of which display the same results, but in two different format of plot:

* The **CA-Procrustes Plot** shows the typical CA plot, whose scores have been moved to fit the reference curve displayed in gray.
* The **Curve Plot** shows the projections of those same points onto the reference curve, from one end to another. Values along the $x$-axis are the indices of points, which are used to derive a ranking of points, from left to right. The distance along the $y$-axis gives the squared Euclidean distance from the point its orthogonal projection on the reference curve.

The `ca_procrustes_ser()` function is run automatically. The Seriation Explorer is an interactive panel, from which the investigator can select points, to re-run and re-fit correspondence analysis. Points are selected using either the plot or the curve plot with the mouse cursor via the `brushedPoints()` function in `shiny`. While the plot is useful for most selections, if the investigator wants to only include points which lie nearest to the reference curve, the curve plot facilitates selection of points within a chosen distance.

When points are selected, their color should automatically change. Note that the selection tool gathers points, and so if a particular shape of a cluster or grouping is to be selected, the brush tool can be used repeatedly to select multiple points. In order to deselect and start over, one double-clicks on the plot.

After selecting points of interest, the investigator then clicks **Recompute with Selection**. Correspondence analysis and Procrustes fitting will then be performed again on a matrix constructed only of the selected row and column elements, omitting any rows and columns which contain all zeros. The plots will automatically refresh. When one believes they have found a well seriated sequence, one clicks on the **Save Seriation as Strand** button. This will log the observed seriation as a `strand` inside a `strands` object (no changes occur to the plots), using the `add_strands()` function.

### Creating Consensus Seriation

After one has selected at least two strands, a consensus seriation can be produced. To do so, click on the **Lakhesize Strands** button in the sidebar, which will perform function `lakhesize()` on the selected strands. Plots will automatically appear in Consensus Seriation and Criteria panels. In the Consensus Seriation panel are included tabs of the biplots of the row and column PCA, along with a matrix plot that gives the size and concentration coefficient of the seriated matrix (not all row and column elements in the input data must be included). In the bottom left Criteria panel, the coefficients of agreement and concentration of each strand are displayed on separate tabs. 

### Modifying Data

#### Deleting Strands

In the process of performing exploratory seriations, it might become clear that one or more strands are poorly seriated, and hence eliminating them might improve the resulting seriated matrix. 

On the Modify panel to the bottom right, one can use the input box labelled **Delete Strand (Indices Will Resort Automatically, Undo Will Re-Add as Last Strand)**, to type in the number of strand to be deleted. Pressing the **Delete** button will remove that strand and automatically readjust the the consensus seriation. The numbering of the strands automatically adjusts. One can add the deleted strand back in, by pressing the **Undo** button. The deleted strand is returned to the end of the strand list, and again the consensus seriation is adjusted automatically.

#### Suppressing Row/Column Elements

There likewise might be a consistent row or column element which appears to be poorly seriated, which the investigator would like to suppress from selection. Following the execution of **Lakhesize Strands**, one can press the **Run Deviance Test** button on the sidebar. The tabs marked "Deviance (Rows)" and "Deviance (Columns)" are now populated with a table of the top ten row and column elements which have the highest $p$~values per the hypothesis test. In the Modify panel, any elements can be entered into the input boxes marked **Select Rows/Columns to Suppress,** either by selecting them from a list or by typing. The Seriation Explorer will automatically remove the row or column elements from the plot. Suppressing row or column elements does not affect previously selected strands. To restore elements, simply delete them from the suppression list in the Modify panel.

### Exporting Results

Exporting results can be done at any time following the command of **Lakhesize Strands**, either in the course of analysis or at the end. The **Export Data** button will download a \code{list} object in \code{rds} format which contains the following objects:

* `$consensus` An object of the `lakhesis` class that contains the information on the consensus seriation (see above on the `lakhesize()` function)
* `$strands` An object the `strands` class, the inputs used to produce the `lakhesis` class object

The default filename is automatically timestamped. The \code{rds} file can then be read back into R for further analysis. 


